### 2020-06-03 serial.tex
Not done but see comment about W.

Both forward and backward perspectives
must yield identical \emph{measurement} (e.g., the length of the incubation
period of a given individual is the same whether measured forward from
the time of infection or backward from the time of symptom onset).
Consequently, if we restrict attention to individuals with a given
incubation period, say $\tau$ days, and we count how many became infected at time $\psymp$
or how many first showed symptoms at time $\ssymp=\psymp+\tau$ then we must obtain
the same number. In general, no matter how delays are distributed, if
$\mathcal P$ and $\mathcal S$ represent the sizes of primary and
secondary cohorts then we must have
\begin{equation}
W(\psymp) \psize(\psymp) f_\psymp(\tau) = \ssize(\ssymp) b_\ssymp(\tau) \,,
\label{eq:match}
\end{equation}
where $W(\psymp)$, the ``weight'' of the primary cohort $\psize(\psymp)$, represents the average number of forward intervals that an individual in cohort $\psize(\psymp)$ produces over the course of their infection.
When we measure within-individual delays, we expect $W(\psymp) \leq 1$ because only a subset of individuals who experience the primary event (e.g., infection) will eventually experience the secondary event (e.g., symptom onset). 
When we measure between-individual delays, we expect $W(\psymp)$ to change throughout an epidemic because individuals infected in the beginning of an epidemic will infect more than one individual, on average, over the course of their infection, whereas those infected later will infect less than one individual, on average.

Substituting $\psymp=\ssymp-\tau$, it follows that
\begin{equation}
b_\ssymp(\tau)= \frac{W(\ssymp-\tau) \psize(\ssymp-\tau) f_{\ssymp-\tau} (\tau)}{\ssize(\ssymp)}\,.
\label{eq:backward}
\end{equation}
If we are considering incubation periods, the left hand side of
this equation is the probability density that an individual
who became symptomatic at time $\ssymp$ had an incubation period of
length $\tau$. 
From the right hand side, we see that this probability density depends on the weight parameter $W(\psymp-\tau)$ (in this case, the proportion of symptomatic infection),
the time-varying primary cohort size at the earlier time $\psize(\ssymp-\tau)$ (in this case, the number of individuals became infected at time $\ssymp-\tau$),
and the forward delay distribution $f_{\ssymp-\tau}(\tau)$ (in this case, the probability density that someone who was infected at time $\ssymp-\tau$ became symptomatic at time $\ssymp$).

@@

Generation and serial intervals determine the time scale of disease transmission, and are therefore critical to dynamical modeling of infectious outbreaks.
Here, we show that the initial \emph{forward} serial-interval distribution --- measured during the exponential growth phase of an epidemic --- provides the correct link between the exponential growth rate $r$ and the reproduction number \RR.
In general, the forward serial-interval distributions will not match the intrinsic or the backward serial-interval distributions.
Further, the mean forward serial interval can decrease over time due to epidemic dynamics. 
Failing to account for these differences can result in underestimation of \RR.

Our study underlines the importance of carefully defining measured epidemiological time distributions.
Previous studies have shown that it matters whether generation
intervals are measured forward or backward \citep{nishiura2010time,champredon2015intrinsic,britton2019estimation};
we generalize these ideas and show that they apply to other epidemiological distributions.
Changes in the backward delay distributions due to changing cohort sizes are expected to be a pervasive feature of outbreak dynamics.
Recent analyses of COVID-19 epidemics have attempted to reconstruct epidemic time series of symptomatic cases or infected cases from the reports of confirmed cases by using time-independent delay distributions from infection or symptom onset to reporting (e.g., \cite{tempvar, park2020potential,scire2020reproductive,shim2020transmission});
these studies effectively assume constant backward delay distributions.
Although such approaches may be able to roughly match the time scale of an epidemic, we recommend using dynamical approaches that model incidence as well as reporting delays explicitly (e.g., \cite{flaxman2020estimating}).

While our results support the use of serial-interval distributions for calculating \RR, 
they also reveal gaps in current practices in incorporating serial-interval distributions into outbreak analyses.
\cite{thompson2019improved} recently emphasized the importance of using up-to-date serial-interval data for accurate estimation of time-varying reproduction numbers.
However, our study demonstrates that early serial-interval data is required to estimate the initial \RR during the exponential growth phase and that using up-to-date serial-interval data (without accounting for changes in the forward serial-interval distribution) can, in fact, underestimate the initial \RR.
Future studies should explore how neglecting changes in the forward serial-interval distribution can affect the estimates of \RR beyond the exponential growth phase and potentially re-assess existing estimates of \RR.

Many studies have already estimated the (time-varying) reproduction numbers for COVID-19 epidemics using aggregated serial-interval data (e.g., \cite{tempvar,du2020serial,pan2020jama,zhang2020evolving});
these studies should also re-assess whether they appropriately considered the \emph{forward} serial interval.
Given large geographical variation in the spread of the COVID-19, the forward serial-interval distributions are also likely to vary across countries.
We suggest that modelers should aim to characterize spatiotemporal variation in forward serial-interval distributions and understand their changes over time.
These modeling approaches should be necessarily coupled with intensive
epidemiological investigation through contact tracing at the very
beginning of the outbreak in order to characterize the initial forward
serial-interval distribution, particularly for an epidemic of an
emerging disease such as COVID-19.

Our study also underlines the fact that the serial-interval distribution depends not only on the generation-interval and incubation-period distributions, but also on their intrinsic correlations.
This implies that not considering this correlation can bias the estimates of the intrinsic generation-interval distribution from the serial-interval distribution.
Furthermore, previous studies that tried to estimate the generation-interval distributions from the observed serial intervals often ignored the dynamical differences between the realized incubation periods of infectors (backward-looking) and those of infectees (forward looking) (e.g., \cite{klinkenberg2011correlation, ganyani2020estimating}).
Better statistical tools for teasing apart the intrinsic generation-interval distributions from the observed serial intervals are needed.

Here, we assume that all individuals develop symptoms and that the entire transmission process, including all relevant epidemiological delays, is known exactly.
In practice, asymptomatic and presymptomatic transmission of COVID-19 makes measuring serial intervals difficult \citep{bai2020presumed,he2020temporal,wei2020presymptomatic}.
Biases in the observed serial intervals will necessarily bias the estimates of \RR. 
Furthermore, serial intervals do not take into account asymptomatic transmission; 
neglecting asymptomatic transmission may also bias the estimates of \RR \citep{park2020time}.

While we focused on the effect of epidemiological dynamics on the serial-interval distributions, other factors, such as intervention strategies, can also affect the generation- and serial-interval distributions.
Individual-level interventions, such as case isolation, and behavioral responses directly affect individuals' ability to transmit and will shorten the forward generation- and serial-intervals.
Population-level interventions, such as social distancing, can squeeze contacts into household and induce clustering, which will in turn shorten the realized generation interval due to local depletion of the susceptible pool within households \citep{park2019inferring}.

>>>>

Both forward and backward perspectives
must yield identical \emph{measurement} (e.g., the length of the incubation
period of a given individual is the same whether measured forward from
the time of infection or backward from the time of symptom onset).
Consequently, if we restrict attention to individuals with a given
incubation period, say $\tau$ days, and we count how many became infected at time $\psymp$
or how many first showed symptoms at time $\ssymp=\psymp+\tau$ then we must obtain
the same number. In general, no matter how delays are distributed, if
$\mathcal P$ and $\mathcal S$ represent the sizes of primary and
secondary cohorts then we must have
\begin{equation}
W(\psymp) \psize(\psymp) f_\psymp(\tau) = \ssize(\ssymp) b_\ssymp(\tau) \,,
\label{eq:match}
\end{equation}
where $W(\psymp)$, the ``weight'' of the primary cohort, represents the average number of forward intervals that an individual in cohort $\psize(\psymp)$ produces over the course of their infection.
When we measure within-individual delays, we expect $W(\psymp) \leq 1$ because only a subset of individuals who experience the primary event (e.g., infection) will eventually experience the secondary event (e.g., symptom onset). 
For between-individual delays, we expect $W(\psymp)$ to change throughout an epidemic, because individuals infected earlier in an epidemic will infect more individuals on average than those infected later.

Substituting $\psymp=\ssymp-\tau$, it follows that
\begin{equation}
b_\ssymp(\tau)= \frac{W(\ssymp-\tau) \psize(\ssymp-\tau) f_{\ssymp-\tau} (\tau)}{\ssize(\ssymp)}\,.
\label{eq:backward}
\end{equation}
If we are considering incubation periods, the left hand side of
this equation is the probability density that an individual
who became symptomatic at time $\ssymp$ had an incubation period of
length $\tau$. 
\jd{I need to keep thinking about what a cohort is for incubation period. Here, you seem to be putting people who don't show symptoms into the forward but not the backward cohort. How sure are you that this is right? The non-symptomatic people also seem to have $\tau$s?}
From the right hand side, we see that this probability density depends on the weight parameter $W(\psymp-\tau)$ (in this case, the proportion of symptomatic infection),
the time-varying primary cohort size at the earlier time $\psize(\ssymp-\tau)$ (in this case, the number of individuals became infected at time $\ssymp-\tau$),
and the forward delay distribution $f_{\ssymp-\tau}(\tau)$ (in this case, the probability density that someone who was infected at time $\ssymp-\tau$ became symptomatic at time $\ssymp$).

@@

Generation and serial intervals determine the time scale of disease transmission, and are therefore critical to dynamical modeling of infectious outbreaks.
Here, we show that the initial \emph{forward} serial-interval distribution --- measured during the exponential growth phase of an epidemic --- provides the correct link between the exponential growth rate $r$ and the reproduction number \RR.
In general, the forward serial-interval distributions will not match the intrinsic or the backward serial-interval distributions.
In particular, the mean forward serial interval can decrease over time due to epidemic dynamics. 
Failing to account for these effects can result in underestimation of \RR.

Our study underlines the importance of carefully defining measured epidemiological time distributions.
Previous studies have shown that it matters whether generation
intervals are measured forward or backward \citep{nishiura2010time,champredon2015intrinsic,britton2019estimation};
we generalize these ideas and show that they apply to other epidemiological distributions.
Changes in the backward delay distributions due to changing cohort sizes are expected to be a pervasive feature of outbreak dynamics.
Recent analyses of COVID-19 epidemics have attempted to reconstruct epidemic time series from by using time-independent delay distributions from infection or symptom onset to reporting (e.g., \cite{tempvar, park2020potential,scire2020reproductive,shim2020transmission});
these studies effectively assume constant backward delay distributions.
Although such approaches may be able to roughly match the time scale of an epidemic, we show here that they are subject to bias. Dynamical approaches that model explicitly model incidence as well as reporting delays (e.g., \cite{flaxman2020estimating}) can avoid this problem, but at the cost of more complexity in estimation. There is a need for simple methods that can complement more explicit approaches of addressing this problem.

While our results support the use of serial-interval distributions for calculating \RR, 
they also reveal gaps in current practices in incorporating serial-interval distributions into outbreak analyses.
\cite{thompson2019improved} recently emphasized the importance of using up-to-date serial-interval data for accurate estimation of time-varying reproduction numbers.
We show here, however, that if changes the forward serial interval through time are not accounted for, using up-to-date serial-interval data can, in fact, exacerbate the underestimation of initial \RR.
Future studies should explore how neglecting changes in the forward serial-interval distribution can affect the estimates of \RR beyond the exponential growth phase and potentially re-assess existing estimates of \RR.

Many studies have already estimated the (time-varying) reproduction numbers for COVID-19 epidemics using aggregated serial-interval data (e.g., \cite{tempvar,du2020serial,pan2020jama,zhang2020evolving});
these studies should also re-assess whether they appropriately considered the \emph{forward} serial interval.
Forward serial-interval distributions are also likely to vary across countries.
We suggest that modelers should aim to characterize spatiotemporal variation in forward serial-interval distributions and understand their changes over time.
These modeling approaches should be coupled with epidemiological investigation through contact tracing. 
Going forward, an additional advantage of early, intensive contact tracing of emerging diseases is that this is the best way to characterize the initial forward serial-interval distribution.

Our study also underlines the fact that the serial-interval distribution depends not only on the generation-interval and incubation-period distributions, but also on their intrinsic correlations.
This implies that not considering this correlation can bias the estimates of the intrinsic generation-interval distribution from the serial-interval distribution.
Previous studies that tried to estimate the generation-interval distributions from the observed serial intervals often ignored the dynamical differences between the realized incubation periods of infectors (backward-looking) and those of infectees (forward looking) (e.g., \cite{klinkenberg2011correlation, ganyani2020estimating}).
Better statistical tools for teasing apart the intrinsic generation-interval distributions from the observed serial intervals are needed.

Here, we assume that all individuals develop symptoms and that the entire transmission process, including all relevant epidemiological delays, is known exactly.
In practice, identifying who infected whom is difficult in general, and asymptomatic and presymptomatic transmission of COVID-19 exacerbates this difficulty \citep{bai2020presumed,he2020temporal,wei2020presymptomatic}.
Biases in the observed serial intervals will necessarily bias the estimates of \RR. 
Furthermore, when one of the individuals is asymptomatic, there is no symptom-based serial interval; 
neglecting the time scale asymptomatic transmission may also bias the estimates of \RR \citep{park2020time}.

While we focus here on the effect of epidemiological dynamics on the serial-interval distributions, other factors, such as intervention strategies, can also affect the generation- and serial-interval distributions.
Individual-level interventions, such as case isolation, and behavioral responses directly affect individuals' ability to transmit and will shorten the forward generation- and serial-intervals.
Population-level interventions, such as social distancing, can squeeze contacts into household and induce clustering, which will in turn shorten the realized generation interval due to local depletion of the susceptible pool within households \citep{park2019inferring}.
